### 2025-07-10 manuscript/part2.tex
intor type system

\section{High-Level Architecture Diagram}
update diagram such that cusomt services also extends the lsp providers
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{graphics/langiumArchitecture.png}
  \caption{Architecture overview of a Langium project}
  \label{fig:langium-architecture}
\end{figure}

@@

Langimus uses validators which check every ast node through visitors if the ... ctype fit. ..d jfl;

\subsection*{Base axioms}
\[
  \begin{array}{rcll}
    \tau   & ::=                                & \mathsf{Bool}
    \;\bigm|\; \mathsf{Int}_{s,w}
    \;\bigm|\; \mathsf{Array}<\mathsf{Int}_{s,w}>
           & (s\in\{\top,\bot\},\;0<w<2^{29}-1)                 \\[1ex]
    \Gamma & ::=                                & \emptyset
    \;\bigm|\; \Gamma,\,x:\tau
  \end{array}
\]
\\
\[
  \frac{}{\Gamma \vdash \mathit{true} : \mathsf{Bool}}
  \; (\textsc{T-True}), \quad
  \frac{}{\Gamma \vdash \mathit{false} : \mathsf{Bool}}
  \; (\textsc{T-False})
\]
\[
  \frac{}
  {\Gamma \;\vdash\; n^{(s,w)} : \mathsf{Int}_{s,w}}
  \; (\textsc{T-IntLit}), \quad
  where\;
  s \in \{\top, \bot\},
  \quad
  w \in \mathbb{N},
  \quad
  1 \;\le\; w \;\le\; 2^{29}-1
\]
\[
  \frac{\vdash\;\mathsf{Int}_{s,w}\;}
  {\vdash\;\mathsf{Array}<\mathsf{Int}_{s,w}>}
  \quad(\textsc{Ty-Array})
\]

\subsection*{Inference rules}
\begin{multicols}{2}

  %–– Column 1 ––%
  \[
    \begin{aligned}
      \frac{x:\tau\in\Gamma}{\Gamma \vdash x : \tau}
       & (\textsc{T-Var})   \\[2ex]
      \frac{\Gamma \vdash e : \tau}{\Gamma \vdash (\,e\,) : \tau}
       & (\textsc{T-Paren}) \\[2ex]
      \frac{\Gamma \vdash e : \mathsf{Bool}}
      {\Gamma \vdash \neg e : \mathsf{Bool}}
       & (\textsc{T-Not})
    \end{aligned}
  \]

@@

  %–– Column 2 ––%
  \[
    \begin{aligned}
      \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s,w}
        \quad
        \Gamma \vdash e_2 : \mathsf{Int}_{s,w}}
      {\Gamma \vdash e_1 < e_2   : \mathsf{Bool}}
       & (\textsc{T-Lt})  \\[1ex]
      \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s,w}
        \quad
        \Gamma \vdash e_2 : \mathsf{Int}_{s,w}}
      {\Gamma \vdash e_1 <= e_2 : \mathsf{Bool}}
       & (\textsc{T-Le})  \\[1ex]
      \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s,w}
        \quad
        \Gamma \vdash e_2 : \mathsf{Int}_{s,w}}
      {\Gamma \vdash e_1 > e_2   : \mathsf{Bool}}
       & (\textsc{T-Gt})  \\[1ex]
      \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s,w}
        \quad
        \Gamma \vdash e_2 : \mathsf{Int}_{s,w}}
      {\Gamma \vdash e_1 >= e_2 : \mathsf{Bool}}
       & (\textsc{T-Ge})  \\[2ex]
      \frac{\Gamma \vdash e_1 : \tau
        \quad
        \Gamma \vdash e_2 : \tau}
      {\Gamma \vdash e_1 == e_2 : \mathsf{Bool}}
       & (\textsc{T-Eq})  \\[1ex]
      \frac{\Gamma \vdash e_1 : \tau
        \quad
        \Gamma \vdash e_2 : \tau}
      {\Gamma \vdash e_1 \:!= e_2 : \mathsf{Bool}}
       & (\textsc{T-Neq}) \\[2ex]
      \frac{\Gamma \vdash e_1 : \mathsf{Bool}
        \quad
        \Gamma \vdash e_2 : \mathsf{Bool}}
      {\Gamma \vdash e_1 \:\&\&\: e_2 : \mathsf{Bool}}
       & (\textsc{T-And}) \\[1ex]
      \frac{\Gamma \vdash e_1 : \mathsf{Bool}
        \quad
        \Gamma \vdash e_2 : \mathsf{Bool}}
      {\Gamma \vdash e_1 \:||\: e_2 : \mathsf{Bool}}
       & (\textsc{T-Or})
    \end{aligned}
  \]
\end{multicols}

% now the large probabilistic-assign rule full width:
\[
  \frac{
    \Gamma \vdash e_1 : \tau
    \quad
    \Gamma \vdash e_2 : \mathsf{Int}_{s,w}
    \quad
    \Gamma \vdash e_3 : \mathsf{Int}_{s,w}
    \quad
    \Gamma \vdash e_4 : \tau
  }
  {\Gamma \vdash e_1 \{\,e_2 :\!:\! e_3\}\, e_4 : \tau}
  \quad(\textsc{T-ProbAssign})
\]

\[
  \begin{aligned}
    \frac{\Gamma\vdash e : \mathsf{Array}\langle\mathsf{Int}_{s,w}\rangle
      \quad
      \Gamma\vdash i : \mathsf{Int}_{s,w}}
    {\Gamma\vdash e[i] : \mathsf{Int}_{s,w}}
     & (\textsc{T-Index})     \\[2ex]
    \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s_1,w_1}
      \quad
      \Gamma \vdash e_2 : \mathsf{Int}_{s_2,w_2}}
    {\Gamma \vdash \mathrm{Bernoulli}(e_1,e_2) : \mathsf{Int}_{\bot,1}}
     & (\textsc{T-Bernoulli}) \\[2ex]
    \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s_1,w_1}
      \quad
      \Gamma \vdash e_2 : \mathsf{Int}_{s_2,w_2}}
    {\Gamma \vdash \mathrm{Uniform}(e_1,e_2) : \mathsf{Int}_{s_2,w_2}}
     & (\textsc{T-Uniform})
  \end{aligned}
\]
\subsection*{Statement Typing Rules}

\[
  \frac{\Gamma\vdash e_1 : \mathsf{Array}<\mathsf{Int}_{s,w}>
  \quad
  \Gamma\vdash i    : \mathsf{Int}_{s,w}
  \quad
  \Gamma\vdash e_2  : \mathsf{Int}_{s,w}}
  {\Gamma\vdash e_1[i] = e_2 \;\mathsf{ok}}
  \quad(\textsc{T-Update})
\]

@@

\[
  \tau \equiv_{comp} \tau_e\quad\text{iff}\quad
  \begin{cases}
    \tau_e = \mathsf{Bool},\ \tau = \mathsf{Bool};                                       \\[4pt]
    \tau_e = \mathsf{Int}_{\top,w_1},\ \tau = \mathsf{Int}_{\top,w_2},\ w_1 \le w_2;     \\[4pt]
    \tau_e = \mathsf{Int}_{\top,w_1},\ \tau = \mathsf{Int}_{\bot,w_2},\ w_1 - 1 \le w_2; \\[4pt]
    \tau_e = \mathsf{Int}_{\bot,w_1},\ \tau = \mathsf{Int}_{\top,w_2},\ w_1 + 1 \le w_2; \\[4pt]
    \tau_e = \mathsf{Int}_{\bot,w_1},\ \tau = \mathsf{Int}_{\bot,w_2},\ w_1 \le w_2;     \\[4pt]
    \mathbf{arrays\;analogous\;to\;int}
  \end{cases}
\]

>>>>

\section{High-Level Architecture Diagram}
update diagram such that cusomt services also extends the lsp providers
\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{graphics/langiumArchitecture.png}
  \caption{Architecture overview of a Langium project}
  \label{fig:langium-architecture}
\end{figure}

@@

Langium offers the option of formalizing semantic checks and type systems through a static validation service which can be registered in its dependency-injection framework.
Once registered, this service traverses the abstract syntax tree via a visitor pattern, invoking node-type specific validation routines at each step.
The following sections describe the implementation of static semantics and the design of the type system.

\subsubsection*{Base axioms}
\[
  \begin{array}{rcll}
    \tau   & ::=                                & \mathsf{Bool}
    \;\bigm|\; \mathsf{Int}_{s,w}
    \;\bigm|\; \mathsf{Array}\langle\mathsf{Int}_{s,w}\rangle
           & (s\in\{\top,\bot\},\;0<w<2^{29}-1)                 \\[1ex]
    \Gamma & ::=                                & \emptyset
    \;\bigm|\; \Gamma,\,x:\tau
  \end{array}
\]
\[
  \frac{}{\Gamma \vdash \mathit{true} : \mathsf{Bool}}
  \; (\textsc{T-True}), \quad
  \frac{}{\Gamma \vdash \mathit{false} : \mathsf{Bool}}
  \; (\textsc{T-False})
\]
\[
  \frac{}
  {\Gamma \;\vdash\; n^{(s,w)} : \mathsf{Int}_{s,w}}
  \; (\textsc{T-IntLit}), \quad
  where\;
  s \in \{\top, \bot\},
  \quad
  w \in \mathbb{N},
  \quad
  1 \;\le\; w \;\le\; 2^{29}-1
\]
\[
  \frac{\vdash\;\mathsf{Int}_{s,w}\;}
  {\vdash\;\mathsf{Array}\langle\mathsf{Int}_{s,w}\rangle}
  \quad(\textsc{Ty-Array})
\]

\subsubsection*{Inference rules}
\begin{multicols}{2}

  \[
    \begin{aligned}
      \frac{x:\tau\in\Gamma}{\Gamma \vdash x : \tau}
       & (\textsc{T-Var})   \\[2ex]
      \frac{\Gamma \vdash e : \tau}{\Gamma \vdash (\,e\,) : \tau}
       & (\textsc{T-Paren}) \\[2ex]
      \frac{\Gamma \vdash e : \mathsf{Bool}}
      {\Gamma \vdash \neg e : \mathsf{Bool}}
       & (\textsc{T-Not})
    \end{aligned}
  \]

@@

  \[
    \begin{aligned}
      \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s,w}
        \quad
        \Gamma \vdash e_2 : \mathsf{Int}_{s,w}}
      {\Gamma \vdash e_1 < e_2   : \mathsf{Bool}}
       & (\textsc{T-Lt})  \\[1ex]
      \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s,w}
        \quad
        \Gamma \vdash e_2 : \mathsf{Int}_{s,w}}
      {\Gamma \vdash e_1 <= e_2 : \mathsf{Bool}}
       & (\textsc{T-Le})  \\[1ex]
      \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s,w}
        \quad
        \Gamma \vdash e_2 : \mathsf{Int}_{s,w}}
      {\Gamma \vdash e_1 > e_2   : \mathsf{Bool}}
       & (\textsc{T-Gt})  \\[1ex]
      \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s,w}
        \quad
        \Gamma \vdash e_2 : \mathsf{Int}_{s,w}}
      {\Gamma \vdash e_1 >= e_2 : \mathsf{Bool}}
       & (\textsc{T-Ge})  \\[2ex]
      \frac{\Gamma \vdash e_1 : \tau
        \quad
        \Gamma \vdash e_2 : \tau}
      {\Gamma \vdash e_1 == e_2 : \mathsf{Bool}}
       & (\textsc{T-Eq})  \\[1ex]
      \frac{\Gamma \vdash e_1 : \tau
        \quad
        \Gamma \vdash e_2 : \tau}
      {\Gamma \vdash e_1 \:!= e_2 : \mathsf{Bool}}
       & (\textsc{T-Neq}) \\[2ex]
      \frac{\Gamma \vdash e_1 : \mathsf{Bool}
        \quad
        \Gamma \vdash e_2 : \mathsf{Bool}}
      {\Gamma \vdash e_1 \:\&\&\: e_2 : \mathsf{Bool}}
       & (\textsc{T-And}) \\[1ex]
      \frac{\Gamma \vdash e_1 : \mathsf{Bool}
        \quad
        \Gamma \vdash e_2 : \mathsf{Bool}}
      {\Gamma \vdash e_1 \:||\: e_2 : \mathsf{Bool}}
       & (\textsc{T-Or})
    \end{aligned}
  \]
\end{multicols}

\[
  \frac{
    \Gamma \vdash e_1 : \tau
    \quad
    \Gamma \vdash e_2 : \mathsf{Int}_{s,w}
    \quad
    \Gamma \vdash e_3 : \mathsf{Int}_{s,w}
    \quad
    \Gamma \vdash e_4 : \tau
  }
  {\Gamma \vdash e_1 \{\,e_2 :\!:\! e_3\}\, e_4 : \tau}
  \quad(\textsc{T-ProbAssign})
\]

\[
  \begin{aligned}
    \frac{\Gamma\vdash e : \mathsf{Array}\langle\mathsf{Int}_{s,w}\rangle
      \quad
      \Gamma\vdash i : \mathsf{Int}_{s,w}}
    {\Gamma\vdash e[i] : \mathsf{Int}_{s,w}}
     & (\textsc{T-Index})     \\[2ex]
    \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s_1,w_1}
      \quad
      \Gamma \vdash e_2 : \mathsf{Int}_{s_2,w_2}}
    {\Gamma \vdash \mathrm{Bernoulli}(e_1,e_2) : \mathsf{Int}_{\bot,1}}
     & (\textsc{T-Bernoulli}) \\[2ex]
    \frac{\Gamma \vdash e_1 : \mathsf{Int}_{s_1,w_1}
      \quad
      \Gamma \vdash e_2 : \mathsf{Int}_{s_2,w_2}}
    {\Gamma \vdash \mathrm{Uniform}(e_1,e_2) : \mathsf{Int}_{s_2,w_2}}
     & (\textsc{T-Uniform})
  \end{aligned}
\]
\subsubsection*{Statement Typing Rules}

\[
  \frac{\Gamma\vdash e_1 : \mathsf{Array}\langle\mathsf{Int}_{s,w}\rangle
  \quad
  \Gamma\vdash i    : \mathsf{Int}_{s,w}
  \quad
  \Gamma\vdash e_2  : \mathsf{Int}_{s,w}}
  {\Gamma\vdash e_1[i] = e_2 \;\mathsf{ok}}
  \quad(\textsc{T-Update})
\]

@@

\[
  \tau \equiv_{comp} e\quad\text{iff}\quad
  \begin{cases}
    \tau_e = \mathsf{Bool},\ \tau = \mathsf{Bool};                                       \\[4pt]
    \tau_e = \mathsf{Int}_{\top,w_1},\ \tau = \mathsf{Int}_{\top,w_2},\ w_1 \le w_2;     \\[4pt]
    \tau_e = \mathsf{Int}_{\top,w_1},\ \tau = \mathsf{Int}_{\bot,w_2},\ w_1 - 1 \le w_2; \\[4pt]
    \tau_e = \mathsf{Int}_{\bot,w_1},\ \tau = \mathsf{Int}_{\top,w_2},\ w_1 + 1 \le w_2; \\[4pt]
    \tau_e = \mathsf{Int}_{\bot,w_1},\ \tau = \mathsf{Int}_{\bot,w_2},\ w_1 \le w_2;     \\[4pt]
    \mathbf{arrays\;analogous\;to\;int}
  \end{cases}
\]
